Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-04-08 endofquarterpaper.qmd

The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]. The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023]. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations between quantitity and diversity of art, and population density increases in other regions and time periods [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992]. Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

@@

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum. This is in accordance with the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.

>>>>

The distribution of objects across sites is highly skewed (@fig-site-map C), with a median number of `r artifact_count_median` objects, but Vogelherd has `r nvogel`. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`), confirming that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is an increase in modularity between the Transitional period and the Proto-Aurignacian, before an decrease into the Early Aurignacian, followed by an increase with the Evolved Aurignacian.

@@

These results are consistent with existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992]. Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of intergroup social networks at around 30 k BP, however, our network analysis shows evidence of intergroup social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted. Dividing the data up into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024]. Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].

@@

Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity. However, these patterns should not be viewed as representing discrete, consistent groups to which all the sites within the group belong, but rather as representing opposite ends of a cultural variant spectrum. We adopt the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups. In this case, restricted sign variability vs. broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.
